Economic Evaluation of Mis-speficied Geographically Weighted Regression Models for Site-specific Nitrogen Management

### 2023-06-24 Writing/manuscript_gwr.rmd

**Abstract**: Geographically weighted regression (GWR) has been presented as a prominent tool to estimate site-specific yield response functions to derive site-specific input rate recommendations from data generated by on-farm precision experimentation (OFPE). However, the current application of GWR in this context assumes quadratic yield response function even though many of the yield-input relationships has been shown to be better represented by the quadratic-plateau functional form. This is due to lack of readily available methods and satistical packages that allow researchers to estimate non-linear model under the GWR framework, including quadratic-plateau. The key objective of this paper is to investigate the possibility of bias in estimating the economic value of site-specific recommendation maps based on GWR with quadratic functional form assumption when in reality the true underlying yield-input relationship is quadratic-plateau. We consider two sets of functional-form specifications used by two recent published studies of site-specific input management that apply GWR to data from OFPE. We show that both GWR approaches suffer significant positive bias, which can lead to large over-estimation of the economic value of applying GWR methods. We reveal that in many circumstances using GWR is far less profitable than claimed by previous studies, which provides a necessary cautionary message to researchers who might otherwise “over-sell” to other researchers the value of using GWR methods in empirical work involving the estimation of site-specific input management. We encourage the statistical community to extend the current GWR capability by allowing researchers to model yield-inpute response more flexibly than currently possible.

@@

On any given field, it is typical for U.S. Midwestern row crop farmers to pay for 1-ha grid soil sampling and chemical analysis every three or four years [@morris2018strengths]. The data are often used to derive fertilizer application recommendations, for example by applying comparatively simple formulas provided by university extension services (e.g., Culman, et al. 2020). But the expense of soil sampling and analysis is driving development of sensor technologies that gather data on site-specific field and crop characteristics or proxies of those characteristics, and the conversion of those data into recommended nitrogen fertilizer (N) application rate maps (usually called "prescriptions" or $R_x$s).^[For example, Trimble Agriculture's GreenSeeker and AgLeader's OptRx® systems use optical sensors mounted on fertilizer application booms to digitally record vegetative index imagery as farm machinery is driven through fields during early crop growth stages. GreenSeeker then uses Oklahoma State University's Sensor-based N Rate Calculator to convert those data into recommended nitrogen fertilizer (N) $R_x$s. Veris Technologies manufactures sensor equipment that can be towed through fields to gather apparent electroconductivity (EC) data, which can be used to infer properties of soil content. The "decision tool" software FieldRevealTM allows users to upload EC data, elevation data, and vegetative index data, which FieldRevealTM's professional agronomists analyze to provide $R_x$s to farmers. FieldApex creates N $R_x$s using foliar vegetative index data from satellite imagery, and SlantRange uses drone imagery for similar purposes.]

However, deploying sophisticated sensors to gather field characteristics data is also costly, and the effects of their generation and use on farming profitability remains in question [@gardner2021economic]. Recognizing the significant costs of obtaining useful, high-resolution soil and field properties data, @trevisan2021spatial demonstrated a method of providing site-specific management recommendations without such data. They estimated site-specific yield response functions by applying GWR techniques on data from OFPEs [@bullock2019data], and reported dramatic results, estimating an average potential gain of \$65 ha$^{-1}$ from applying seed and N at optimal uniform rates (instead of at the rates the farmers would have chosen if not participating in the research), and an additional potential gain of \$58 ha$^{-1}$ when changing from optimal uniform application rates to optimal site-specific application rates. This is rather surprising given the often-reported poor statistical accuracy of GWR estimates [@griffith2008spatial; @paez2011simulation; @da2016multiple; @harris2019simulation] and also much lower value of variable rate application (VRA) over uniform rate application (URA) reported in previous studies (cite). @rakshit2020 also applied GWR to on-farm experiment data. However, while they did estimate site-specific yield response functions, they did not derive site-specific economically optimal nitrogen rates based on them, and thus did not conduct economic analysis, unlike @trevisan2021spatial. @evans2020assessment explored using GWR with a linear model specification to detect low- and high-response areas, but like @rakshit2020, they also did not estimate site-specific economically optimal nitrogen rates.

While the results reported by @trevisan2021spatial look rosy and GWR looks like a promising method, there are two potentially dangerous factors that may have caused them to significantly over-estimate the value of GWR as a tool to estimate site-specific EONR. First, they used quadratic functional form in their GWR estimation even though it is widely accepted by scientific community that corn yield response to nitrogen is better approximated by quadratic-plateau functional form. It is likely that their choice of quadratic functional form is merely a compromise because there did (does) not exist any statistical software of package that implements GWR with quadratic-plateau functional form.

The second source of over-estimation bias is endemic to any study that evaluates the value of variable rate application, not just @trevisan2021spatial. In general, researchers take the following steps to estimate the value of VRA using OFPE data [e.g., @anselin2004spatial; @swinton1998evaluating; @bullock2002adding; @liu2006site]: (1) estimate site-specific yield response functions based on a method chosen by the researcher; (2) use the estimated yield response functions to estimate the economically optimal site-specific input rates (EOIRs) and economically optimal uniform input rates; (3) plug the estimated site-specific EOIRs and uniform EOIR into the estimated yield response functions to calculate estimated profits under both strategies; and (4) find the difference in the two strategies' estimated profits. However, this approach can result in significant bias in the estimation of the value of VRA because of the step 3's assumption that the estimated yield response functions are the true yield response functions. @trevisan2021spatial provided an example at step (3) by assuming that their GWR estimates provided "true" site-specific yield response functions. This assumption is generally unavoidable for researchers since they cannot observe any real field's true site-specific yield response functions. But, inaccurate estimation of yield response function in this methodology can result in significant bias in the estimation of the value of VRA. This bias problem can be particularly severe when the method used tends to exaggerate the true heterogeneity in yield response functions, such as GWR [@czarnota2015evaluating;].

The primary objective of this study is examine the degree and direction of bias in the estimation of the economic value of GWR-based site-specific nitrogen rate recommendations (or "variable rate application", VRA) when GWR model is specified as quadratic when the true yield response function follows the quadratic-plateau functional form. We evaluate two sets of GWR quadratic specifications used in @rakshit2020 and @trevisan2021spatial. Both of them assumed a quadratic response of yield to nitrogen rate ($N$). However, while @trevisan2021spatial estimated the coefficient on $N^2$ as a global parameter, @rakshit2020 estimated it locally for each point. At first, it appears that the specification used by @rakshit2020 is superior since it is more flexible than the specification used by @trevisan2021spatial. However, the flexibility can be a disadvantage rather than an advantage because the estimation of economically optimal nitrogen rate (EONR) can be highly sensitive to the coefficient estimates on $N^2$, and GWR often provides inaccurate local estimation of that coefficient. It is critical to keep in mind that it is not our objective to measure the bias in the coefficient estimation by GWR, unlike [@griffith2008spatial; @paez2011simulation; @da2016multiple; @harris2019simulation]. Rather, we are interested in the economic implications of the statistical inaccuracy in estimating the coefficients of the model in this study. As part of identifying bias in the estimation of the economic value of VRA based on the GWR approaches, we evaluate their actual values. Finally, we also evaluate their relative performances compared to uniform rate application (URA) based on shape-constrained additive model approach. It should be noted that the economical optimization in this study only includes farmers' private profit, while the environmental externalities (such as fertilizer runoff and leaching) are not considered.

@@

An extensive literature examines the choice of functional form in yield response estimation [@cerrato1990comparison, @frank1990comparison, @bullock1994quadratic, @tembo2008crop]. Our simulation assumes that each cell $(j, k)$'s "true" yield response function is assumed to follow the commonly-used quadratic-plateau functional form:

>>>>

**Abstract**: Geographically weighted regression (GWR) has been presented as a prominent tool to estimate site-specific yield response functions to derive site-specific input rate recommendations from data generated by on-farm precision experimentation (OFPE). However, the current application of GWR in this context assumes quadratic yield response function even though many of the yield-input relationships has been shown to be better represented by the quadratic-plateau functional form. This is due to lack of readily available methods and statistical packages that allow researchers to estimate non-linear model under the GWR framework, including quadratic-plateau. The key objective of this paper is to investigate the possibility of bias in estimating the economic value of site-specific recommendation maps based on GWR with quadratic functional form assumption when in reality the true underlying yield-input relationship is quadratic-plateau. We consider two sets of functional-form specifications used by two recent published studies of site-specific input management that apply GWR to data from OFPE. We show that both GWR approaches suffer significant positive bias, which can lead to large over-estimation of the economic value of applying GWR methods. We reveal that in many circumstances using GWR is far less profitable than claimed by previous studies, which provides a necessary cautionary message to researchers who might otherwise “over-sell” to other researchers the value of using GWR methods in empirical work involving the estimation of site-specific input management. We encourage the statistical community to extend the current GWR capability by allowing researchers to model yield-input response more flexibly than currently possible.

@@

On any given field, it is typical for U.S. Midwestern row crop farmers to pay for 1-ha grid soil sampling and chemical analysis every three or four years [@morris2018strengths]. The data are often used to derive fertilizer application recommendations, for example by applying comparatively simple formulas provided by university extension services [e.g., @culman2020tri]. But the expense of soil sampling and analysis is driving development of sensor technologies that gather data on site-specific field and crop characteristics or proxies of those characteristics, and the conversion of those data into recommended nitrogen fertilizer (N) application rate maps (usually called "prescriptions" or $R_x$s).^[For example, Trimble Agriculture's GreenSeeker and AgLeader's OptRx® systems use optical sensors mounted on fertilizer application booms to digitally record vegetative index imagery as farm machinery is driven through fields during early crop growth stages. GreenSeeker then uses Oklahoma State University's Sensor-based N Rate Calculator to convert those data into recommended nitrogen fertilizer (N) $R_x$s. Veris Technologies manufactures sensor equipment that can be towed through fields to gather apparent electroconductivity (EC) data, which can be used to infer properties of soil content. The "decision tool" software FieldRevealTM allows users to upload EC data, elevation data, and vegetative index data, which FieldRevealTM's professional agronomists analyze to provide $R_x$s to farmers. FieldApex creates N $R_x$s using foliar vegetative index data from satellite imagery, and SlantRange uses drone imagery for similar purposes.]

However, deploying sophisticated sensors to gather field characteristics data is also costly, and the effects of their generation and use on farming profitability remains in question [@gardner2021economic]. Recognizing the significant costs of obtaining useful, high-resolution soil and field properties data, @trevisan2021spatial demonstrated a method of providing site-specific management recommendations without such data. They estimated site-specific yield response functions by applying GWR techniques on data from OFPEs [@bullock2019data], and reported dramatic results, estimating an average potential gain of \$65 ha$^{-1}$ from applying seed and N at optimal uniform rates (instead of at the rates the farmers would have chosen if not participating in the research), and an additional potential gain of \$58 ha$^{-1}$ when changing from optimal uniform application rates to optimal site-specific application rates. This is rather surprising given the often-reported poor statistical accuracy of GWR estimates [@griffith2008spatial; @paez2011simulation; @da2016multiple; @harris2019simulation] and also much lower value of variable rate application (VRA) over uniform rate application (URA) reported in previous studies [@bullock1998does; @thrikawala1999economic; @koch2004economic]. @rakshit2020 also applied GWR to on-farm experiment data. However, while they did estimate site-specific yield response functions, they did not derive site-specific economically optimal nitrogen rates based on them, and thus did not conduct economic analysis, unlike @trevisan2021spatial. @evans2020assessment explored using GWR with a linear model specification to detect low- and high-response areas, but like @rakshit2020, they also did not estimate site-specific economically optimal nitrogen rates.

While the results reported by @trevisan2021spatial look rosy and GWR looks like a promising method, there are two potentially dangerous factors that may have caused them to significantly over-estimate the value of GWR as a tool to estimate site-specific EONR. First, they used quadratic functional form in their GWR estimation even though it is widely accepted by scientific community that corn yield response to nitrogen is better approximated by quadratic-plateau functional form. It is likely that their choice of quadratic functional form is merely a compromise because there did (does) not exist any statistical software of package that implements GWR with quadratic-plateau functional form. 
The second source of over-estimation bias is endemic to any study that evaluates the value of variable rate application, not just @trevisan2021spatial. In general, researchers take the following steps to estimate the value of VRA using OFPE data [e.g., @anselin2004spatial; @swinton1998evaluating; @bullock2002adding; @liu2006site]: (1) estimate site-specific yield response functions based on a method chosen by the researcher; (2) use the estimated yield response functions to estimate the economically optimal site-specific input rates (EOIRs) and economically optimal uniform input rates; (3) plug the estimated site-specific EOIRs and uniform EOIR into the estimated yield response functions to calculate estimated profits under both strategies; and (4) find the difference in the two strategies' estimated profits. However, this approach can result in significant bias in the estimation of the value of VRA because of the step 3's assumption that the estimated yield response functions are the true yield response functions. @trevisan2021spatial provided an example at step (3) by assuming that their GWR estimates provided "true" site-specific yield response functions. This assumption is generally unavoidable for researchers since they cannot observe any real field's true site-specific yield response functions. But, inaccurate estimation of yield response function in this methodology can result in significant bias in the estimation of the value of VRA. This bias problem can be particularly severe when the method used tends to exaggerate the true heterogeneity in yield response functions, such as GWR [@czarnota2015evaluating].

The primary objective of this study is examine the degree and direction of bias in the estimation of the economic value of GWR-based site-specific nitrogen rate recommendations (i.e., variable rate application of nitrogen) when GWR model is specified as quadratic when the true yield response function follows the quadratic-plateau functional form. We evaluate two sets of GWR quadratic specifications used in @rakshit2020 and @trevisan2021spatial. Both of them assumed a quadratic response of yield to nitrogen rate ($N$). However, while @trevisan2021spatial estimated the coefficient on $N^2$ as a global parameter, @rakshit2020 estimated it locally for each point. At first, it appears that the specification used by @rakshit2020 is superior since it is more flexible than the specification used by @trevisan2021spatial. However, the flexibility can be a disadvantage rather than an advantage because the estimation of economically optimal nitrogen rate (EONR) can be highly sensitive to the coefficient estimates on $N^2$, and GWR often provides inaccurate local estimation of that coefficient. It is critical to keep in mind that it is not our objective to measure the bias in the coefficient estimation by GWR, unlike [@griffith2008spatial; @paez2011simulation; @da2016multiple; @harris2019simulation]. Rather, we are interested in the economic implications of the statistical inaccuracy in estimating the coefficients of the model in this study. As part of identifying bias in the estimation of the economic value of VRA based on the GWR approaches, we evaluate their actual values. Finally, we also evaluate their relative performances compared to uniform rate application (URA) based on shape-constrained additive model approach. It should be noted that the economical optimization in this study only involves farmers' private profit, while the environmental externalities (such as fertilizer runoff and leaching) are not considered.

@@

An extensive literature examines the choice of functional form in yield response estimation [@cerrato1990comparison, @frank1990comparison, @bullock1994quadratic, @tembo2008crop]. Our simulation assumes that each cell $(j, i)$'s "true" yield response function is assumed to follow the commonly-used quadratic-plateau functional form:
